---
title: "R tips March 2025"
author: "Nikita M."
format:
  html:
    page-layout: article
    toc: true
    html-math-method: katex
    highlight: tango
date: "2025-03-13"
title-block-banner: false
categories: [R-tips]
---


```{r setup, include=FALSE}
library(lobstr)
# options(warnPartialMatchDollar = TRUE)
```

## Floating-Point Precision in R

Floating-point numbers in R are subject to precision limitations due to [IEEE 754](https://en.wikipedia.org/wiki/IEEE_754) representation.
This section explores how seemingly simple operations can lead to unexpected results.

### **Example 1: Floating-Point Comparison Fails**

```{r floating-point-compare, echo=TRUE}
.1 == .3 / 3
```

**Why does this return `FALSE`?**

-   Due to how floating-point arithmetic works in computers, numbers like `0.1`, `0.3`, and `0.3 / 3` are stored in binary.

-   Some decimal fractions **cannot be represented exactly in binary**, leading to tiny rounding errors.

-   Even though mathematically `.1` and `.3 / 3` should be the same, their internal representations **are not exactly equal**.

### **Example 2: Inspecting the Actual Value Stored in Memory**

```{r floating-point-print, echo=TRUE}
.3 / 3
print(.3 / 3, digits=16)
```

**Explanation:**

-   The first output shows what R normally prints, which is rounded for display.

-   The second output uses `digits=16` to show more of the **actual stored precision**.

-   This reveals that `0.3 / 3` is actually stored as `0.09999999999999999`, **not exactly** `0.1`.

### **Example 3: Using `all.equal()` Instead of `==`**

```{r floating-point-all-equal, echo=TRUE}
all.equal(.1, .3 / 3)
```

**Why does this return `TRUE`?**

-   `all.equal()` does not check for **exact** equality but allows a **small tolerance** to account for floating-point errors.

-   It is the **preferred way** to compare floating-point numbers in R.

### **Example 4: Machine Epsilon & Adjusting Tolerance**

```{r floating-point-eps, echo=TRUE}
.Machine$double.eps
all.equal(.1, .3 / 3, tolerance = 2.2e-17)
```

**Explanation:**

-   `.Machine$double.eps` is the smallest possible difference that R can distinguish between two floating-point numbers.

-   The default tolerance of `all.equal()` is `sqrt(.Machine$double.eps)`, but we can **adjust it manually** if needed.

📖 **More details:** [The R Inferno](https://www.burns-stat.com/documents/books/the-r-inferno/)

------------------------------------------------------------------------

## Memory Allocation in R

Memory management in R determines how objects are stored and how much space they occupy.
**Understanding how R allocates memory** can help optimize performance, especially with large datasets.

### **Example 5: Comparing Memory Usage of Integer and Double**

```{r memory-allocation, echo=TRUE}
x <- 5    # Stored as double
obj_size(x)  
```

```{r memory-allocation2, echo=TRUE}
y <- 5L   # Stored as integer
obj_size(y) 
```

**Why are both `x` and `y` 56 bytes?**

-   Even though `y` is an **integer (4 bytes)** and `x` is a **double (8 bytes)**, R adds metadata overhead.

-   Every object in R carries around **48 bytes of metadata**, including type and reference count.

-   Memory is aligned in **multiples of 8 bytes**, so the total allocation rounds up to 56 bytes.

### **Example 6: Memory Usage for Large Vectors**

```{r large-vector-memory, echo=TRUE}
x_vec <- rep(5, 1000000)   # Double vector
y_vec <- rep(5L, 1000000)  # Integer vector

obj_size(x_vec)  # ~8 MB (each number is 8 bytes)
obj_size(y_vec)  # ~4 MB (each number is 4 bytes)
```

**Why does the integer vector use half the memory?**

-   Each **integer** (`5L`) takes **4 bytes per element**.

-   Each **double** (`5`) takes **8 bytes per element**.

-   For **large datasets**, using integers instead of doubles can **save significant memory**.

### **Example 7: Type Conversion Affects Memory Usage**

```{r memory-conversion, echo=TRUE}
y_vec[2] = 5.0
obj_size(y_vec)  # Changes due to type conversion

new_vec = lapply(y_vec, as.integer)
obj_size(new_vec) 

new_vec[2] = 5.5
new_vec[1:3]
obj_size(new_vec) 
```

**Explanation:**

-   Assigning `5.0` to `y_vec` **forces** R to convert the entire vector to **double**, increasing memory usage.

-   `lapply(y_vec, as.integer)` converts the vector back to integers, reducing its size.

-   But assigning `5.5` back **forces conversion to double again**.

------------------------------------------------------------------------

## Partial Matching in R

R allows **partial matching** of list elements, which can lead to unintended behaviour and difficult-to-debug errors.

### **Example 8: Partial Matching Works, But It’s Risky**

```{r partial-matching, echo=TRUE}
test_function <- function(alpha = NULL, power = NULL, p = NULL, p0 = NULL) {
  user_defined <- list(alpha=alpha, power=power, p=p, p0=p0)
  user_defined <- Filter(Negate(is.null), user_defined)
  return(user_defined)
}

results = test_function(alpha=0.05, power=0.8)
print(str(results))
print(results$p)  # Partial matching
```

**Why is `results$p` returning `NULL`?**

-   Because `p` was not provided in `test_function()`, but R still allows partial matching.

### **Example 9: Unexpected Partial Matching in Lists**

```{r unexpected-partial-match, echo=TRUE}
x = list(data = 2, power = 10)
x$p  # Partial matching: returns 10

if (x$p == 10) cat("variable p == 10\n")

# Avoiding partial matching
options(warnPartialMatchDollar = TRUE)
print(x$p)  # Now warns about partial matching

print(x[["p"]])  # Explicit retrieval, avoiding unexpected matches
```

**Best Practices:**

-   Use **`options(warnPartialMatchDollar = TRUE)`** to catch unintended partial matches.

-   Always use **`x[["p"]]`** instead of `x$p` to ensure explicit matching.

📖 **More details:** [Advanced R - Subsetting](https://adv-r.hadley.nz/subsetting.html?q=partial#section-1)

------------------------------------------------------------------------

```{r}
#| include: false
options(warnPartialMatchDollar = FALSE)
```
